### 2012-07-10 2.tex
Writing on chapters 2, 4

A naïve account of this contrast is based on the assumption that segments or timing units must either be parsed into prosodic structures like the syllable, or be subject to phonological repair. Early precedents for this can be found in dissertations by \citet[10f.]{Hooper1973} and \citet[57f.]{Kahn1976}, and it is a fundamental component of many later frameworks (e.g., \citealt{Ito1989a}, \citealt{Noske1992}, \citealt{OT}). In English, [bl] is a permissible onset, but unlike some languages (e.g., Moroccan Arabic \emph{bni} `building', \emph{bnat} `daughters', \emph{bniqa} `closet'), [bn] is not. \citet[][19f.]{Wolf2009} suggest, for instance, that an underlying /bnɪk/ would be realized as [nɪk]; it is also possible to imagine resolution by prothesis---[əbnɪk]---or anapytxis---[bənɪk], and indeed \citet{Davidson2006b} finds that English speakers use all three of these repairs when asked to mimic foreign pronunciations of obstruent-nasal onsets like /bn/.

@@

This chapter focuses on the hypothesis, implicit in \citeauthor{Shademan2006}'s statement, that there is an inherent gradient or continuum between possible and impossible words, against an null hypothesis that the observed patterns of judgements are little more than an artifact of the wordlikeness task itself. \S\ref{gradhyp} places this debate in a historical perspective, and argues that the naïve but widely accepted form of this hypothesis fails the test of falsifiability. In \S\ref{2evaluation}, a quantitative evaluation of a revised version of this hypothesis reveals a poor match between the gradience in speakers' judgements and contemporary computational models of wordlikeness. Taken together, the findings argue against the gradience hypothesis.

% What \citet[32]{Cohn2006} calls ``gradient well-formedness''

\section{The gradience hypothesis} \label{gradhyp}

@@

\subsubsection{\emph{The sound pattern of English}}

In \emph{The sound pattern of English} (SPE), \citet{SPE} argue that a complete solution to wordlikeness must countenance gradience. In addition to the contrast of \emph{blick} and \emph{bnick}, they introduce a third nonce word \emph{bnzk}, which they take to be even less well-formed than ``impossible'' \emph{bnick}.

@@

Despite the considerable attention given to the proposals of \emph{SPE} in the wake of that book's publication in 1968, the \emph{SPE} wordlikeness model has received almost no attention in the literature. At the risk of explaining what might be no more than an an accidental gap in the historical record, the novel aspects of the \emph{SPE} model---gradience derived from similarity to existing lexical entries---seem to have been overshadowed by compelling arguments against \citeauthor{SPE}'s assumption (see also \citealt{Halle1962}, \citealt{Stanley1967}) that wordlikeness contrasts derive from properties of underlying forms. \citet{Shibatani1973} observes that there are some generalizations about surface forms which give rise to wordlikeness contrasts, but cannot be stated as constraints on underlying forms. An example from German is shown in (\ref{fd}) below.

@@

\noindent The plural appears to preserve a contrast in final obstruent voicing which is absent in the singular:\footnote{While there is a contentious debate as to whether devoicing in German is completely neutralizing (e.g., \citealt{Fourakis1984}) or not \citep[e.g.,][]{Port1985}, it is irrelevant to the discussion at hand.} word-final voiced stops are never found in German, and \citet[95]{Shibatani1973} reports that ``it is easy to show that a native speaker of German rejects those forms \emph{on the ground that they end in voiced obstruents}'' (emphasis in original). However, given that root-final voicing is not predictable (e.g., /ɡʀaːt \alt{} ɡʀaːd/), the process of \textsc{Final Devoicing} cannot be any kind of lexical redundancy and is mysterious under the \emph{SPE} account. The alternative proposed by \citeauthor{Shibatani1973} and by \citet{Clayton1976}, is that to say that nonce words like [ɡʀaːd] are ill-formed not beacuse of any underlying property, but because they fail to undergo an otherwise-exceptionless phonological process, \textsc{Final Devoicing}, or equivalently, the surface-true generalization it derives.

@@

The arguments of \citeauthor{Shibatani1973} and others led theorists to focus their attention on properties of surface representations as determinants of wordlikeness. Though syllabification plays no role in \emph{SPE}, the syllable is discussed in many earlier studies, and it received particular attention in the 1970s. \citet{Hooper1973} and \citet{Kahn1976} argue that the syllable is useful for defining wordlikeness generalizations.\footnote{\citet{Steriade1999} and \citet{Blevins2003}, however, argue that a number of phonotactic generalizations previously stated in syllabic terms can be reanalyzed without making reference to syllables.} \citeauthor{Hooper1973} argues, for instance, that [bn], impossible as an English onset, is unobjectionable as a syllable contact cluster in nonce words like \emph{stabnik} (or in names like \emph{Abner}), and that this demonstrates the superiority of syllable-based wordlikeness generalizations. This already signals further trouble for alternative accounts which focus on underlying forms. Syllabification may span morphs, is generally predictable, and is universally non-contrastive, and as a consequence, few posit in to be present in underlying representations \citep[though see, e.g.,][]{Vaux2003}. \citeauthor{Hooper1973} also points to loanword adaptations which produce native syllable structure \citep[e.g.,][]{Carlisle1991} as evidence that syllabification is part of the phonological computation.

Further enrichments to the theory are provided by the autosegmental theory of the syllable \citep{McCarthy1979b}, which envisions the syllable as an articulated tree structure \citep[as envisioned by][]{Pike1947a}, and theories like prosodic licensing \citep{Ito1989a}, in which syllabification triggers phonological repairs.

@@

Using the head-term preference paradigm, \citet{Jusczyk1993b} and \citet{Friederici1993} find that typically-developing children as young as 9 months of age distinguish between nonce words which are and are not phonotactically valid in their target language. \citet{Jusczyk1994} report that 9-month-old children acquiring English also show preferences for nonce words with high positional probability over those with low positional probability. Faciliatory effects of positional probability (i.e., shorter latencies) are reported for other nonce word tasks conducted with adults, including single-word shadowing \citep{Vitevitch1997,Vitevitch1998}, same/different judgements (\citealt{Vitevitch1999a}, \citealt{Luce2001}, \citealt{Lipinski2005}, \citealt{Vitevitch2005}), and lexical decision \citep{Pylkkanen2002a}.

%\citet{Mattys1999,Mattys2001}
%\citet{McQueen1998}

The most interesting result of these studies comes from the first experiment of \citealt{Vitevitch1997}, which finds that speakers' wordlikeness ratings of bisyllabic words are correlated with the positional probabilities of the constituent syllables. Unfortunately, neither \citeauthor{Vitevitch1997} nor any of the other aforementioned psycholinguists make any effort to eliminate the possibility that the low positional probability stimuli are ``impossible'' words of English. In fact, in Chapter \ref{clusters}, it is argued that many of the stimuli used by \citet{Vitevitch1997}, including even the high probability stimuli, contain illicit word-medial consonant clusters. While \citeauthor{Vitevitch1997} neither control nor manipulate the well-formedness of medial clusters, in a post-hoc test they consider a probabilistic measure of cluster well-formedness to a post-hoc test, which reveals that cluster well-formedness is correlated with syllable-internal positional probabilities and the wordlikeness judgements. \citeauthor{Vitevitch1997} ultimately conclude that it cannot explain all the variation in wordlikeness.

The aforementioned studies all conclude that the gradient measure of positional probability correlates with behavioral results. As the flaws of the \citet{Vitevitch1997} study demonstrate, the aforementioned studies do little to tease apart the gradient and categorical aspects of phonotactics. More generally, they do little to distinguish between positional probability and closely correlated measures like bigram probability (see \S\ref{bigram} below) or neighborhood density, since these studies carefully select stimuli which either have high or low values for all of positional probability, bigram probablility, and neighborhood density. This is particularly troublesome given that positional probability was created \emph{ex nihilo}; in contrast, the effects of neighborhood density in various psycholinguistic tasks are emergent properties of many models of speech production (e.g., \citealt{Luce1998}, \citealt{Luce2000}) and perception (e.g., \citealt{Marslen-Wilson1984,Marslen-Wilson1987}, \citealt{McClelland1986}, \citealt{Norris1994}, \citealt{Norris2000}).

\subsection{A naïve gradience axiom}

Recent critiques of this syllabification-based model focus on the existence of intermediate wordlikeness ratings (see also \citealt{Coleman1997}, \citealt{Anttila2008}).

@@

The observation itself cannot be questioned, but \citeauthor{Hayes2008a} and \citeauthor{Albright2009a} do not explicitly state why this data is relevant to the construction of models of wordlikeness. Presumably, these authors believe that these patterns of judgements demonstrate that wordlikeness, as an internal state, is gradient simply because subjects make use of intermediate degrees of wordlikeness in judgement tasks. This proposition, generalized below, is ``naïve'' not because it lacks sophistication, but because it is rooted in a belief in naïve realism, a philosophy which holds that perception provides a relatively direct picture of the nature of the world, and in general, an influential view in the cognitive sciences (see \citealt{Fodor1981a} for a critique).

\begin{unlabeledexample}
\textsc{Naïve Gradience Proposition}: If graded judgements of a concept or mental state use intermediate ratings, that concept or mental state is gradient
\end{unlabeledexample}

This proposition is an indicative conditional. While the antecedent of this condition, the presence or absence of intermediate ratings in a graded judgement task, is readily observable, it has no bearing on the truth or falsity of the conditional proposition. However, whereas a false consequent (e.g., an all-or-nothing concept) would potentially falsify the proposition, the consequent, a mental state, cannot be directly observed. If this is the case, then both the proposition and the consequent (the proposition that wordlikeness is itself gradient) fail the test of falsification, placing it beyond the reach of scientific inquiry.

@@

\citeauthor{Armstrong1983} do not view their results as evidence that subjects have an internal model of odd numbers which is at odds with the formal, all-or-nothing definition, and in fact, several additional experiments show that subjects deploy the extension of the formal definition in other tasks. Reviewing this study, \citet[215]{Schutze2011} writes that the experiments show that judgements are ``sensitive to factors other than our underlying competence''. In the case of wordlikeness, a case can be made that one such factor is similarity to existing words (e.g., \citealt[][151, fn. 27]{LSLT}, \citealt{Greenberg1964}, \citealt{Ohala1986b}, \citealt{Bailey2001}, \citealt{Schutze2005}). However, the mere fact that subjects use intermediate ratings does not show that they do so with any systematicity, or that, e.g., the contrast between \emph{447} and \emph{7} with respect could be given any satisfying explanation. \citeauthor{Schutze2011} further suggests that gradience responses is a property of tasks, not concepts.

@@

The remainder of this chapter is devoted to evaluating the gradience hypothesis, using state-of-the-art computational wordlikeness models and data from English.

@@

The evaluation makes exclusive use of previously published English wordlikeness data. For inclusion in the evaluation, a study must conform to all three of the following conditions. First, the stimuli must consist only of monosyllabic words presented auditorily. Secondly, a significant portion of the stimuli must contain gross phonotactic violations (e.g., \emph{bnick}). Finally, the ratings, averaged across subjects, must be publicly available. The large-scale studies by \citet{Bailey2001} and \citet{Shademan2006,Shademan2007} are ineligible because the former lack stimuli with gross phonotactic violations and the latter data are not available to the public in any form. All data are reproduced in Appendix \ref{appendixA}

@@

\citet{Scholes1966} conducted a number of English wordlikeness judgement with middle school children. The data used here come from his ``experiment 5'', in which 63 monosyllabic items were presented to 33 seventh-grade students. For each stimulus, the subjects produced forced choice ``yes''/``no'' answers to the question of whether the item ``is likely to be usable as a word of English''. \citet{Hayes2008a} and \citet{Albright2009a} analyze this data as gradient by performing an item averaging using the fraction of ``yes'' answers for each stimulus (see also \citealt{Pierrehumbert1994}, \citealt{Coleman1997}, \citealt{Frisch2000}). For instance, 22 of the 33 students answered ``yes'' for \emph{shlerk} [ʃlɚk], so it is assigned a score of $0.666$.\footnote{This procedure conflates intraspeaker variation, which in some cases is considerable \citep{Shademan2007} with speaker-internal gradience, and its adoption here should not be construed as an endorsement. Rather, it is provided solely for comparison with prior studies using the \citeauthor{Scholes1966} data.} These stimuli are all ostensibly non-words, but include \emph{clung} [klʌŋ], the preterite and past participle of the verb \emph{cling}, and \emph{brung} [brʌŋ], a dialectical past participle for \emph{bring}. These two words were excluded and the remaining 61 stimuli were submitted to analysis.

@@

The four models used here consist of a binary baseline and three computationally implemented gradient models. These three models are chosen because prior studies have shown they are correlated with wordlikeness judgements. Non-parametric rank correlation statistics are used evaluate the correlation between model predictions and wordlikeness ratings. Rank correlation are used rather than the parametric Pearson correlation statistic that is used by \citet{Hayes2008a}, for instance, because they make none of the potentially troublesome assumptions of the latter method (see \citealt{Albright2009a}:23, fn. 12). The Spearman $\rho$ is most widely known rank correlation statistic, but it is difficult to give a natural interpretation to this quantity. On the other hand, the Kendall $\tau_b$ and Goodman-Kruskal $\gamma$ can be interpreted as fractions of the number of \emph{concordant} and \emph{discordant} pairs \citep{Noether1981}. Consider the case here, in which model scores are compared with wordlikeness ratings. If a model rates \emph{dresp} [dɹɛsp] more wordlike than *\emph{srest} [sɹɛst] and if speakers further rate \emph{dresp} more English-like than \emph{srest}, than \emph{dresp}/\emph{srest} are a concordant pair. If, however, a model and speakers disagree on the relative ranking of \emph{dresp} and \emph{srest}, the pair is discordant. The $\tau_b$ and $\gamma$ differ only in their treatment of ``ties'' (i.e., if e.g., *\emph{dresp} and *\emph{srest} are scored the same, or rated the same); $\tau_b$ includes a correction for ties, whereas $\gamma$ ignores tied pairs. Much like the familiar Pearson correlation, $\rho$, $\tau_b$ and $\gamma$ are all in the range [$-1$, $1$]. Correlations of the four models are given in Table \ref{cor}.

\begin{table} \centering
\begin{tabular}{l l l l l}
\toprule
Spearman $\rho$          & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {0.845}          & {0.765} & {\textbf{0.863}} & {0.648}          \\
\citealt{Scholes1966}    & {0.791}          & {0.762} & {\textbf{0.827}} & {0.827}          \\
\citealt{Albright2003b}  & {0.725}          & {0.429} & {0.708}          & {\textbf{0.742}} \\
\midrule
Kendall $\tau_b$         & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {\textbf{0.716}} & {0.585} & {0.692}          & {0.462}          \\
\citealt{Scholes1966}    & {\textbf{0.664}} & {0.597} & {0.652}          & {0.565}          \\
\citealt{Albright2003b}  & {\textbf{0.599}} & {0.343} & {0.506}          & {0.556}          \\
\midrule
Goodman-Kruskal $\gamma$ & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {\textbf{1.000}} & {0.684} & {0.692}          & {0.462}          \\
\citealt{Scholes1966}    & {\textbf{0.995}} & {0.634} & {0.667}          & {0.614}          \\
\citealt{Albright2003b}  & {\textbf{0.953}} & {0.656} & {0.509}          & {0.575}          \\
\bottomrule
\end{tabular}
\caption{Rank correlations between wordlikeness ratings and phonotactic models surprisingly reveal that the binary baseline meeds or exceeds the coverage of three state-of-the-art phonotactic models. All correlations are significant at $p = 0.05$.}
\label{cor}
\end{table}

@@

The binary baseline used here is a crude implementation of the null hypothesis that there are no gradient effects in wordlikeness judgements. To create such a baseline, it is necessary to distinguish between nonce words which contain a gross phonotactic violation and those which do not. As all stimuli here are monosyllables, this task can be further simplified by separately considering the two major subcomponents of the syllable, the onset and rime. Speakers are particularly adept as separating onsets from rimes (\citealt{Treiman1986,Treiman1995}, \citealt{Fowler1993}), and a large portion of phonotactic violations can be localized to one or the other unit (e.g., \citealt{Fudge1969}, \citealt{Treiman2000}).

The baseline considers a nonce syllable to be well-formed if it consists of both an attested onset and an attested rime; the free combination of these two components is the only mechanism by which this model can generalize beyond attested words. It is surely the case that this is an ultimately insufficient model: \citet{Albright2009a} observes, for instance, that [ɛsp] is a well-formed rime, even though it is found in no word of English. This problem is more acute in languages with more permissive syllable structures than those of English, and thus a sparser lexicon with regard to phonotactic constraints. For instance, \citet{Fischer-Jorgensen1952} and \citet{Vogt1954} assert that there are many accidental gaps (i.e., possible but unattested structures) in the inventory of consonant clusters in Georgian, a language which admits as many as five adjacent consonants.\footnote{Chapter \ref{clusters} reports a similar result for English syllable contact clusters.} Despite these defects, the baseline outperforms the gradient models in most contexts.

@@

For $\tau_b$ and $\gamma$, the interpretation of this result is clear. The gradient models assign rankings to the sets of phonotactically valid and invalid clusters, respectively. For instance, the bigram model favors \emph{troog} [tɹuːg] over \emph{swach} [swætʃ], though neither contains any gross phonotactic violation. Similarly, the bigram model favors \emph{choop} [tʃuːp] over \emph{zhrick} [ʒɹɪk], even though neither word is a possible word of English. However, the majority of such predicted contrasts are not reflected in speakers' judgements; for instance, \emph{troog} is rated less English-like than \emph{swach} \citep{Greenberg1964}, contrary to the model predictions. This is stark evidence against the gradience hypothesis.

@@

This chapter

has evaluated the axiom of gradience as a falsifiable alternative hypothesis.

>>>>

A naïve account of this contrast is based on the assumption that segments or timing units must either be parsed into prosodic structures like the syllable, or be subject to phonological repair. Early precedents for this can be found in dissertations by \citet[10f.]{Hooper1973} and \citet[57f.]{Kahn1976}, and it is a fundamental component of many later frameworks \citep[e.g.,][]{Ito1989a,Noske1992,OT}.

In English, [bl] is a permissible onset, but unlike some languages (e.g., Moroccan Arabic \emph{bni} `building', \emph{bnat} `daughters', \emph{bniqa} `closet'), [bn] is not. \citet[][19f.]{Wolf2009} suggest, for instance, that an underlying /bnɪk/ would be realized as [nɪk]; it is also possible to imagine resolution by prothesis---[əbnɪk]---or anapytxis---[bənɪk], and indeed \citet{Davidson2006b} finds that English speakers use all three of these repairs when asked to mimic foreign pronunciations of obstruent-nasal onsets like /bn/.

@@

This chapter focuses on the hypothesis, implicit in \citeauthor{Shademan2006}'s statement, that there is an inherent gradient or continuum between possible and impossible words, against an null hypothesis that the observed patterns of judgements are little more than an artifact of the wordlikeness task itself. \S\ref{2background} places this debate in a historical perspective, and argues that the naïve but widely accepted form of this hypothesis fails the test of falsifiability. In \S\ref{2evaluation}, a quantitative evaluation of a revised version of this hypothesis reveals a poor match between the gradience in speakers' judgements and contemporary computational models of wordlikeness. Taken together, the findings argue against the gradience hypothesis.

\section{Background} \label{2background}

@@

\subsubsection{\emph{The sound pattern of English}} \label{2spe}

In \emph{The sound pattern of English} (Henceforth \emph{SPE}), \citet{SPE} argue that a complete solution to wordlikeness must countenance gradience. In addition to the contrast of \emph{blick} and \emph{bnick}, they introduce a third nonce word \emph{bnzk}, which they take to be even less well-formed than ``impossible'' \emph{bnick}.

@@

Despite the considerable attention given to the proposals of \emph{SPE} in the wake of that book's publication in 1968, the \emph{SPE} wordlikeness model has received almost no attention in the literature. At the risk of explaining what might be no more than an an accidental gap in the historical record, the novel aspects of the \emph{SPE} model---gradience derived from similarity to existing lexical entries---seem to have been overshadowed by compelling arguments against \citeauthor{SPE}'s assumption \citep[see also][]{Halle1962,Stanley1967} that wordlikeness contrasts derive from properties of underlying forms. \citet{Shibatani1973} observes that there are some generalizations about surface forms which give rise to wordlikeness contrasts, but cannot be stated as constraints on underlying forms. An example from German is shown in (\ref{fd}) below.

@@

\noindent The plural appears to preserve a contrast in final obstruent voicing which is absent in the singular:\footnote{While there is a contentious debate as to whether devoicing in German is completely neutralizing \citep[e.g.,][]{Fourakis1984}) or not \citep[e.g.,][]{Port1985}, it is irrelevant to the discussion at hand.} word-final voiced stops are never found in German, and \citet[95]{Shibatani1973} reports that ``it is easy to show that a native speaker of German rejects those forms \emph{on the ground that they end in voiced obstruents}'' (emphasis in original). However, given that root-final voicing is not predictable (e.g., /ɡʀaːt \alt{} ɡʀaːd/), the process of \textsc{Final Devoicing} cannot be any kind of lexical redundancy and is mysterious under the \emph{SPE} account. The alternative proposed by \citeauthor{Shibatani1973} and by \citet{Clayton1976}, is that to say that nonce words like [ɡʀaːd] are ill-formed not beacuse of any underlying property, but because they fail to undergo an otherwise-exceptionless phonological process, \textsc{Final Devoicing}, or equivalently, the surface-true generalization it derives.

@@

The arguments of \citeauthor{Shibatani1973} and others led theorists to focus their attention on properties of surface representations as determinants of wordlikeness. Though syllabification plays no role in \emph{SPE}, it is crucial to many earlier studies \citep[see][]{Goldsmith2011b}, and it received particular attention in the 1970s. \citet{Hooper1973} and \citet{Kahn1976} argue that the syllable is useful for defining wordlikeness generalizations.\footnote{\citet{Steriade1999} and \citet{Blevins2003}, however, argue that a number of phonotactic generalizations previously stated in syllabic terms can be reanalyzed without making reference to syllables.} \citeauthor{Hooper1973} argues, for instance, that [bn], impossible as an English onset, is unobjectionable as a syllable contact cluster in nonce words like \emph{stabnik} (or in names like \emph{Abner}), and that this demonstrates the superiority of syllable-based wordlikeness generalizations. This already signals further trouble for alternative accounts which focus on underlying forms. Syllabification may span morphs, is generally predictable, and is universally non-contrastive, and as a consequence, few posit in to be present in underlying representations \citep[though see, e.g.,][]{Vaux2003}. \citeauthor{Hooper1973} also points to loanword adaptations which produce native syllable structure \citep[e.g.,][]{Carlisle1991} as evidence that syllabification is part of the phonological computation. Further enrichments to the theory are provided by the autosegmental theory of the syllable \citep{McCarthy1979b}, which envisions the syllable as an articulated tree structure \citep[as envisioned by][]{Pike1947a}, and theories like prosodic licensing \citep{Ito1989a}, in which syllabification triggers phonological repairs.

@@

Using the head-term preference paradigm, \citet{Jusczyk1993b} and \citet{Friederici1993} find that typically-developing children as young as 9 months of age distinguish between nonce words which are and are not phonotactically valid in their target language. \citet{Jusczyk1994} report that 9-month-old children acquiring English also show preferences for nonce words with high positional probability over those with low positional probability. Faciliatory effects of positional probability (i.e., shorter latencies) are reported for other nonce word tasks conducted with adults, including single-word shadowing \citep{Vitevitch1997,Vitevitch1998}, same/different judgements \citep{Vitevitch1999a,Luce2001,Lipinski2005,Vitevitch2005}, and lexical decision \citep{Pylkkanen2002a}.

Most relevant to the question at hand, \citet{Vitevitch1997} and \citet{Frisch2000} find that speakers' wordlikeness ratings of multisyllabic words are correlated wtih the positional probailities of the constituent syllables. Unfortunately, none of these researchers make any effort to eliminate the possibility that the low positional probability stimuli are ``impossible'' words of English. In fact, 
Chapter \ref{clusters} argues
%the author has argued elsewhere \citep{Gorman2012c}
that many of the stimuli used by \citet{Vitevitch1997} and \citet{Frisch2000} contain illicit word-medial consonant clusters. While \citeauthor{Vitevitch1997} neither control nor manipulate the well-formedness of medial clusters, in a post-hoc test they consider a probabilistic measure of cluster well-formedness, which reveals that cluster well-formedness is correlated with syllable-internal positional probabilities and the wordlikeness judgements. \citeauthor{Vitevitch1997} ultimately conclude that it cannot explain all the variation in wordlikeness.

The aforementioned studies all conclude that the gradient measure of positional probability correlates with behavioral results. As the flaws of the \citet{Vitevitch1997} study demonstrate, the aforementioned studies do little to tease apart the gradient and categorical aspects of phonotactics. More generally, they do little to distinguish between positional probability and closely correlated measures like bigram probability (see \S\ref{bigram} below) or neighborhood density, since these studies carefully select stimuli which either have high or low values for all of positional probability, bigram probablility, and neighborhood density. This is particularly troublesome given that positional probability was created \emph{ex nihilo}; in contrast, the effects of neighborhood density in various psycholinguistic tasks are emergent properties of many models of speech production \citep[e.g.,][]{Luce1998,Luce2000} and perception \citep{Marslen-Wilson1984,Marslen-Wilson1987,McClelland1986,Norris1994,Norris2000}.

\subsection{A naïve gradience hypothesis}

Recent critiques of this syllabification-based model focus on the existence of intermediate wordlikeness ratings \citep[see also][]{Coleman1997,Anttila2008}.

@@

As discussed in \S\ref{2spe} above, this is also the view of \citeauthor{SPE} in \emph{SPE}. The observation itself cannot be questioned, but \citeauthor{SPE}, \citeauthor{Hayes2008a}, and \citeauthor{Albright2009a} do not explicitly state why this data is relevant to the construction of models of wordlikeness. Presumably, these authors believe that these patterns of judgements demonstrate that wordlikeness, as an internal state, is gradient simply because subjects make use of intermediate degrees of wordlikeness in judgement tasks. This proposition, generalized below, is ``naïve'' not because it lacks sophistication, but because it is rooted in a belief in naïve realism, a philosophy which holds that perception provides a relatively direct picture of the nature of the world, an influential view in the cognitive sciences in general (see \citealt{Fodor1981a} for a critique).

\begin{unlabeledexample}
\textsc{Naïve Gradience Hypothesis}: If graded judgements of a concept or mental state use intermediate ratings, that concept or mental state is gradient
\end{unlabeledexample}

This proposition has the form of an indicative conditional. While the antecedent of this condition, the presence or absence of intermediate ratings in a graded judgement task, is readily observable, it has no bearing on the truth or falsity of the conditional proposition. However, whereas a false consequent (e.g., an all-or-nothing concept) would potentially falsify the proposition, the consequent, a mental state, cannot be directly observed. If this is the case, then both the proposition and the consequent (the proposition that wordlikeness is itself gradient) fail the test of falsification, placing it beyond the reach of scientific inquiry.

@@

\citeauthor{Armstrong1983} do not view their results as evidence that subjects have an internal model of odd numbers which is at odds with the formal, all-or-nothing definition, and in fact, several additional experiments show that subjects deploy the extension of the formal definition in other tasks. Reviewing this study, \citet[215]{Schutze2011} writes that the experiments show that judgements are ``sensitive to factors other than our underlying competence''. In the case of wordlikeness, a case can be made that one such factor is similarity to existing words (e.g., \citealt[][151, fn. 27]{LSLT}, \citealt{Greenberg1964}, \citealt{Vitz1973}, \citealt{Ohala1986b}, \citealt{Sendlmeier1987}, \citealt{Bailey2001}, \citealt{Schutze2005}). However, the mere fact that subjects use intermediate ratings does not show that they do so with any systematicity, or that, e.g., the contrast between \emph{447} and \emph{7} with respect could be given any satisfying explanation. \citeauthor{Schutze2011} further suggests that gradience responses is a property of tasks, not concepts.

@@

The remainder of this chapter is devoted to testing the revised hypothesis. This is accomplished by an evaluation of computational models of wordlikeness in which categorical and gradient models are placed on an equal footing.

@@

The evaluation makes exclusive use of previously published English wordlikeness data. For inclusion in the evaluation, a study must conform to all three of the following conditions. First, the stimuli must consist only of monosyllabic words presented auditorily. Secondly, a significant portion of the stimuli must contain gross phonotactic violations (e.g., \emph{bnick}). Finally, the ratings, averaged across subjects, must be publicly available. The large-scale studies by \citet{Bailey2001} and \citet{Shademan2006,Shademan2007} are ineligible because the former lack stimuli with gross phonotactic violations and the latter data are not available to the public in any form. All data are reproduced in Appendix \ref{appendixA}.

@@

\citet{Scholes1966} conducted a number of English wordlikeness judgement with middle school children. The data used here come from his ``experiment 5'', in which 63 monosyllabic items were presented to 33 seventh-grade students. For each stimulus, the subjects produced forced choice ``yes''/``no'' answers to the question of whether the item ``is likely to be usable as a word of English''. \citet{Hayes2008a} and \citet{Albright2009a} analyze this data as gradient by performing an item averaging using the fraction of ``yes'' answers for each stimulus \citep[see also][]{Pierrehumbert1994,Coleman1997,Frisch2000}. For instance, 22 of the 33 students answered ``yes'' for \emph{shlerk} [ʃlɚk], so it is assigned a score of $0.666$.\footnote{This procedure conflates intraspeaker variation, which in some cases is considerable \citep{Shademan2007} with speaker-internal gradience, and its adoption here should not be construed as an endorsement. Rather, it is provided solely for comparison with prior studies using the \citeauthor{Scholes1966} data.} These stimuli are all ostensibly non-words, but include \emph{clung} [klʌŋ], the preterite and past participle of the verb \emph{cling}, and \emph{brung} [brʌŋ], a dialectical past participle for \emph{bring}. These two words were excluded and the remaining 61 stimuli were submitted to analysis.

@@

The four models used here consist of a binary baseline and three computationally implemented gradient models. These three models are chosen because prior studies have shown they are correlated with wordlikeness judgements. Non-parametric rank correlation statistics are used evaluate the correlation between model predictions and wordlikeness ratings. Rank correlation are used rather than the parametric Pearson correlation statistic that is used by \citet{Hayes2008a}, for instance, because they make none of the potentially troublesome assumptions of the latter method \citep[see][23, fn. 12]{Albright2009a}.

The Spearman $\rho$ is most widely known rank correlation statistic, but it is difficult to give a natural interpretation to this quantity. On the other hand, the Kendall $\tau_b$ and Goodman-Kruskal $\gamma$ can be interpreted as fractions of the number of \emph{concordant} and \emph{discordant} pairs \citep{Noether1981}. Consider the case here, in which model scores are compared with wordlikeness ratings. If a model rates \emph{dresp} [dɹɛsp] more wordlike than *\emph{srest} [sɹɛst] and if speakers further rate \emph{dresp} more English-like than \emph{srest}, than \emph{dresp}/\emph{srest} are a concordant pair. If, however, a model and speakers disagree on the relative ranking of \emph{dresp} and \emph{srest}, the pair is discordant. The $\tau_b$ and $\gamma$ differ only in their treatment of ``ties'' (i.e., if e.g., *\emph{dresp} and *\emph{srest} are scored the same, or rated the same); $\tau_b$ includes a correction for ties, whereas $\gamma$ ignores tied pairs. Much like the familiar Pearson correlation, $\rho$, $\tau_b$ and $\gamma$ are all in the range [$-1$, $1$]. Correlations of the four models are given in Table \ref{cor}.

\begin{table} \centering
\begin{tabular}{l l l l l}
\toprule
Spearman $\rho$          & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {0.845}          & {0.765} & {\textbf{0.863}} & {0.648}          \\
\citealt{Scholes1966}    & {0.791}          & {0.762} & {\textbf{0.827}} & {\textbf{0.827}} \\
\citealt{Albright2003b}  & {0.725}          & {0.429} & {0.708}          & {\textbf{0.742}} \\
\midrule
Kendall $\tau_b$         & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {\textbf{0.716}} & {0.585} & {0.692}          & {0.462}          \\
\citealt{Scholes1966}    & {\textbf{0.664}} & {0.597} & {0.652}          & {0.565}          \\
\citealt{Albright2003b}  & {\textbf{0.599}} & {0.343} & {0.506}          & {0.556}          \\
\midrule
Goodman-Kruskal $\gamma$ & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {\textbf{1.000}} & {0.684} & {0.692}          & {0.462}          \\
\citealt{Scholes1966}    & {\textbf{0.995}} & {0.634} & {0.667}          & {0.614}          \\
\citealt{Albright2003b}  & {\textbf{0.953}} & {0.656} & {0.509}          & {0.575}          \\
\bottomrule
\end{tabular}
\caption{Rank correlations between wordlikeness ratings and phonotactic models surprisingly reveal that the binary baseline meeds or exceeds the coverage of three state-of-the-art phonotactic models. All correlations are significant at $p = 0.05$.}
\label{cor}
\end{table}

@@

The binary baseline used here is a crude implementation of the null hypothesis that there are no gradient effects in wordlikeness judgements. To create such a baseline, it is necessary to distinguish between nonce words which contain a gross phonotactic violation and those which do not. As all stimuli here are monosyllables, this task can be further simplified by separately considering the two major subcomponents of the syllable, the onset and rime. Speakers are particularly adept as separating onsets from rimes \citep{Treiman1986,Treiman1995,Fowler1993}, and a large portion of phonotactic generalizations and violations can be localized to either onset or rime \citep[e.g.,][]{Fudge1969,Treiman1995,Kessler1997,Treiman2000}.

The baseline considers a nonce syllable to be well-formed if it consists of both an attested onset and an attested rime; the free combination of these two components is the only mechanism by which this model can generalize beyond attested words. It is surely the case that this is an ultimately insufficient model: \citet{Albright2009a} observes, for instance, that [ɛsp] is a well-formed rime, even though it is found in no word of English. This problem is more acute in languages with more permissive syllable structures than those of English, and thus a sparser lexicon with regard to phonotactic constraints. For instance, \citet{Fischer-Jorgensen1952} and \citet{Vogt1954} assert that there are many accidental gaps (i.e., possible but unattested structures) in the inventory of consonant clusters in Georgian, a language which admits as many as five adjacent consonants.\footnote{
%Chapter \ref{clusters} 
\citet{Gorman2012c}
reports a similar result for English syllable contact clusters.} Despite these defects, the baseline outperforms the gradient models in most contexts.

@@

For $\tau_b$ and $\gamma$, the interpretation of this result is clear. The gradient models assign rankings to the sets of phonotactically valid and invalid clusters, respectively. For instance, the bigram model favors \emph{troog} [tɹuːɡ] over \emph{swach} [swætʃ], though neither contains any gross phonotactic violation. Similarly, the bigram model favors \emph{chwoop} [tʃwuːp] over \emph{zhrick} [ʒɹɪk], even though neither word is a possible word of English. However, the majority of such predicted contrasts are not reflected in speakers' judgements; for instance, \emph{troog} is rated less English-like than \emph{swach} \citep{Greenberg1964}, contrary to the model predictions. This is stark evidence against the gradience hypothesis.

@@

\subsection{The gradience hypothesis}

This chapter has evaluated the axiom of gradience as a falsifiable alternative hypothesis. The surprising result is that virtually all of the apparent coverage of state-of-the-art gradient phonotactic models is simply a reflection of their ability to distinguish between the possible and the totally impossible; beyond this, they are unreliable. A trivial baseline, endowed with few abilities to project beyond the observed data, generally outperforms the state of the art. It follows that the projections made by the state-of-the-art gradient models are not like those made by speakers. What remains to be seen is whether any model can be put forth which accurately predicts these intermediate ratings, and whether such a model deploys linguistic representations.
